% langs/apl
\par\begin{footnotesize}\begin{alltt}
Contents of README files:\par
Thank you for your letter of 27 February. The original purpose
of our APL font and macros was to solve the immediate problem
of writing documents with APL code. Therefore, they are free for
anyone who finds them useful. Some people have requested the font
and we have sent them the files appended below (with CMAPL10 and
two conversion programs for downloading pixel files, not given here).
Further, we have nothing against giving our net address to the
users because it would be useful to hear about extensions made
to the system.

The files APLDEF.TEX and TUGBOAT.TEX are identical with the second
versions sent to you (via BITNET) except for minimal changes (there
are some corrections and extensions in APLDEF.TEX, and in
TUGBOAT.TEX one APL line was corrected.) We think the files are
in a "final" form (at least we are not planning any changes).

Our original purpose was to publish the second version, but the only
difference was really the removal of the tilde space character (this
resulted in more complex 1-character macros, but the user cannot
see any difference; moreover, TeX runs only about 14% slower). Thus,
we feel that there is no reason to publish a list of errata in
TUGboat, because any user would inevitably have to use the new
version (the old one has not been released).

We are adding a file containing sample examples on the use of the
APL mnemonics.

Aarno Hohti
\end{alltt}\end{footnotesize}\par
